Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework

### 2016-07-05 ms/ms.Rnw
removed pesky typo where ht occured after figure
Attempt at new introduction

\title{Unifying intraspecific and interspecific variation in tropical tree mortality}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{James S Camac}
\author[a]{Mark Westoby} 
\author[b]{S Joseph Wright}
\author[a]{Daniel S Falster}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Add something significant here}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.C. designed and performed the analysis; D.F. conceived the idea, supervized the development of this work; S.J.W \& M.W. provided biological insight. All authors contributed to the writing of this manuscript.}

@@

\begin{abstract}
Death is an integral process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these are yet to be integrated within a general predictive framework. Here we estimate instantenous mortality rates of 203 tropical rainforest tree species at Barro Colorado Island (BCI), Panama, by combining classic survival analysis techniques with a cross-validated hierarchical Bayesian framework that integrates individual-level, species-level, and plot-level effects. We show that greatest predictive capacity is achieved when when models simultaneously estimate growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth and account for wood density effects on growth-independent terms. Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the largest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

<<preload, echo=FALSE>>=
n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))

@

@@

\dropcap{P}lant mortality is often modelled as a function of recent growth because it reflects the product of many biotic and abiotic factors on a plants overall carbon budget and vigour \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}. As such, growth rates are considered a highly integrated individual-based predictor of plant mortality \cite{Hawkes:2000ib}. Within a species, the assumption is that fast growing individuals (i.e. those that have larger carbon budgets) are more able to buffer stress and therefore have higher survivorship relative to slow growing (i.e. low carbon budget) individuals \cite{Hawkes:2000ib, Keane:2001db}. Empirical studies examining this growth\textendash mortality relationship have broadly supported this assumption, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Consequently, most dynamic vegetation models, including tree gap models \cite{Keane:2001db}, individual-based forest simulators \cite[e.g. SORTIE][]{Pacala:1996vm} and Dynamic Global Vegetation Models \cite[DGVMs; ][]{Woodward:2004ft} use some measure of growth coupled with a stochastic constant (i.e. baseline hazard or minimum chance of death) as the fundamental determinants of plant mortality \cite{Hawkes:2000ib, McDowell:2011dr}.

However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms with overly simplistic assumptions \cite{Hawkes:2000ib, Wyckoff:2000un, McDowell:2011dr}. One of the strongest assumptions made is that both stochastic hazards and the growth-mortality relationship are constant across species \cite{Pacala:1996vm, Hawkes:2000ib, Wyckoff:2000un}. That is, all species have the same random chance of death and exhibit the same tolerance to low growth. This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \cite{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}. Furthermore, because no systematic analysis has been conducted to determine what growth measure best predicts plant mortality, the growth-mortality relationship is often parameterized using a variety growth measures; two of the most common being rates of DBH or basal area growth.

Recently, research has focused on examining inter-specific variation by correlating intrinsic species' growth rates with population mortality rates \cite[e.g.][]{Poorter:2008iu, Wright:2010fl}. In contrast to intraspecific work, these studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species \cite{Loehle:1996vw, Poorter:2008iu, Wright:2010fl}. As such, species appear to exhibit a trade off between greater growth versus greater survival. Ecologists have attempted to explain this interspecific variation by using a variety of functional traits \textemdash attributes of species that are thought to influence vital rates (i.e. survival, growth and reproduction) and ultimately fitness \cite{Westoby:2002tb, Ackerly:2003eb, Falster:2015un}.

Wood density, the biomass invested per unit wood volume, is one such functional trait that has consistently been shown to be negatively correlated with tropical tree mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}. The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \cite{Augspurger:1984wx}. As such, it is hypothesized that species with higher wood density are likely to exhibit lower stochastic hazards. However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-mortality relationship. For example, species with high wood density have been found to be more tolerant to both shade \cite{vanGelder:2006exa} and neighbor competition \cite{Kunstler:2016km}. Studies have also shown that wood density may reduce drought induced carbon starvation \cite{McDowell:2011dr} and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack \cite{Augspurger:1984wx, McDowell:2011dr}. Consequently, species with high wood density may be more tolerant to both stochastic events such as breakage as well as be more tolerant to periods of low growth.

What both intraspecific and interspecific studies have highlighted is that both individual (e.g. growth rates) and species (e.g. function traits) characteristics are likely to be required in order to develop general and accruate models of plant mortality. However, to date, there has been little crossover of ideas between intraspecific and interspecific studies of plant mortality. For example, we are aware of only one study that uses individual based plant mortality empirical models to examine trait effects on baseline hazards \cite{AubryKientz:2013dg}, and none that examine how traits mediate the growth-mortality relationship.

\begin{figure}[ht]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology. A) Our data consist of repeat measures on stem diameter ($D$) and status ($S$, alive or dead) for individual tress at specific census dates ($t_1$, $t_2$, $t_3$). B) We consider three alternative mortality functions, combining growth-independent and growth-dependent effects. For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$. C) The predictive skill of alternative models is evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). We then refit each alternative model 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data. We then average the log loss $\mathcal{L_i}$ for each of these 10 model fits, to give an overall average log loss for each model $\bar{\mathcal{L}}$.}
\label{fig1}
\end{figure}

In this study, we reconcile both intraspecific and interspecific mortality theory by combining classic survival analysis techniques with a cross-validated hierarchical Bayesian framework. Specifically, our framework  instantaneous mortality rates for 203 tropical rainforest trees at Barro Colorado Island (BCI) by integrating individual-level, species-level, and plot-level effects and simultaneously estimating growth-dependent and growth-independent mortality processes. Using this framework we 
identify the most parsimonious model of tropical tree instantaneous mortality rates. We also ask: 1) which  growth measure (i.e. diameter increment growth or area growth)best predicts growth-dependent mortality; 2) Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both? 3) What is the proportion of species variability explained by wood density?

\section*{Framework for modelling mortality}

In classic survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right),
\end{equation}
where $\lambda_i(t)$ is the instantaneous mortality rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$. The challenge then is to identify suitable forms of the instantaneous hazard (mortality) rate function.

We examine the performance of three functional forms of $\lambda_i(t)$ to predict observed patterns of tropical tree mortality:
\begin{align}
\lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} +
\overbrace{\gamma.}^{\text{Baseline hazard}} \label{add_haz}
\end{align}
The simplest model (eq. \ref{base_haz}), assumes a single baseline mortality rate, which is constant irrespective of a plant's growth rate. The second model (eq. \ref{growth_haz}) describes a growth-dependent mortality rate which assumes plants can only die due to a growth related cause. In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached. This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}. The third model (eq. \ref{add_haz}) -- potentially the most biologically realistic -- is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike eq. \ref{growth_haz}, will asymptote at some baseline rate instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). The parameters of eq. \ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality rate $\lambda$ at low annual growth rate. $\beta$ represents the exponential decay rate or the sensitivity of mortality rates to change with growth rate; and $\gamma$, is the asymptote that defines the baseline hazard. $\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters. By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g. windfall, fire) that can kill a plant irrespective of its growth rate.

The central aim of this study is to identify the most parsimonious combination of $\lambda$, growth measure (diameter increment growth or area growth) and wood density that best predicts tropical tree mortality. To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison. Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz} \textendash \ref{add_haz} using either annual diameter increment or area growth rates. In this stage each model was fitted without allowing parameters to vary by species. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure. Following this, we used the best model of stage 1 but allowed its parameters to vary by species. This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance. Because species responses varied considerably, we undertook a third stage of model fitting and selection that explicitly examined whether wood density (a species level trait) explained some of this interspecific variability. This involved using the best model observed in stage 1 and adding wood density effects to each parameter. Finally we then fitted the best wood density combination model with species random effects to the entire dataset. This model was then used to examine mortality curves across wood density gradients. In order to determine the proportion of interspecific variation explained by wood density on each parameter it occurred on we reran the final model form with just random effects on parameters and calculated the percent reduction in unexplained species variation.

At each stage, we evaluated the predictive accuracy of alternative models or predictor variables using K-fold cross-validation (Fig. \ref{fig1}). This involves randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, while ensuring that each K-fold had at least one observation per species. We then fitted models to nine folds and predicted to the tenth fold. This was repeated until each fold had been used nine times for model fitting, and once as a held-out dataset for predicting. Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \cite{Hooten:2015, Stone:1997tg}.

@@

\begin{figure}[ht]
\includegraphics{figures/fig4.pdf}[ht]
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig4}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics{figures/fig5.pdf}[ht]
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig5}
\end{figure}

\begin{figure}[ht]
\includegraphics{figures/fig6.pdf}[ht]
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig6}
\end{figure}

@@

>>>>

\title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{James S Camac}
\author[a,2]{Mark Westoby}
\author[a,3]{Daniel S Falster}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}

@@

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.C. designed and performed the analysis; D.F. conceived the idea, supervized the development of this work; M.W. provided biological insight. All authors contributed to the writing of this manuscript.}

@@

% 250 words
\begin{abstract}
Tree death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework. Here we apply classic techniques from survival analysis within a hierarchical Bayesian and cross-validation framework to derive instantaneous mortality functions for tropical rain forest trees at Barro Colorado Island (BCI) Panama, integrating individual-level, species-level, and plot-level effects. In reaching this point, we show how mortality models including separate growth-dependent and growth-independent components, using stem diameter growth rather than basal-area growth, and attributing the effects of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

<<preload, echo=FALSE>>=
n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
@

@@

\dropcap{P}lant mortality rates are known to vary among individuals within a species, among coexisting species within the same forest, from year-to-year, and between forests [refs]. Current estimates suggest the range of variation is large, e.g. from xxx xxx among plants within the same forest, and from XX-XXX among forests within any given year, or between years. This variation has considerable consequences for forest structure and dynamics. Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools, while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun. In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates \cite{Johnson-2016}. Improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.

Two substantial challenges exist when studying the mortality process in trees. The first is that very large population sizes and/or very long periods of observation are needed for effective inference. This requirement arises is in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$ \cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are needed, e.g. 100s-1000s records. As such, detailed studying of mortality for forest trees has only recently become possible, with the accumulation of data from repeat surveys in long-term monitoring plots containing 1000's trees and spanning several decades \cite{Anderson-Teixeira-2015a}.

\begin{figure}[ht]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) Our data consist of repeat measures on stem diameter ($D$) and status ($S$, alive or dead) for individual tress at specific census dates ($t_1$, $t_2$, $t_3$).
B) We consider three alternative mortality functions, combining growth-independent and growth-dependent effects. The parameters of the model are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes ingrowth rate; and $\gamma$ is the asymptote, or baseline hazard. Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g. windfall, fire) that kill a plant irrespective of its growth rate. For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
C) Log-loss function
D) The predictive skill of alternative models is evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data. We then average the log loss $\mathcal{L_i}$ for each of these 10 model fits, to give an overall average log loss for each model $\bar{\mathcal{L}}$.}
\label{fig1}
\end{figure}

The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate, traits, and size (XX). Generally, however, these influences have not been formalised as hazard functions. Yet such knowledge is needed for predicting mortality into the future and in models of ecosystem function (XX). For projecting the outcomes of a population, hazard functions are invaluable because they ....

A particular challenge is to integrate the apparently conflicting relationships of mortality to growth observed within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species also having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality into a unified model, including individual-level, species-level, and plot-level effects, within a cross-validated hierarchical Bayesian framework for 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. To derive this unified model, we compare the skill of three alternative functional forms for $\lambda_i(t)$ in predicting patterns of tropical tree mortality in a diverse tropical forest in Panama (Fig. \ref{fig1}B). The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases. Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (refs). However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific hypotheses:
\begin{enumerate}\itemsep 0pt
  \item Does mortality vary substantially between censuses?
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of total inter-specific variability explained by wood density?
\end{enumerate}
Answering these questions involves comparing predictive skill of alternative models. To achieve this, we used K-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the five hypotheses evaluated above, and gradually increasing model complexity. At each stage, the best model from the previous stage was taken as input for the next stage. for example, stage 2 involved comparing predictive performance of the three $\lambda$ models described in (Fig. \ref{fig1}B) using either annual diameter increment or area growth rates. In this stage each model was fitted without allowing parameters to vary by species. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure. Following this, we used the best model of stage 2 but allowed its parameters to vary by species. This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance.

@@

\begin{figure}[ht]
\includegraphics{figures/fig4.pdf}
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig4}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics{figures/fig5.pdf}
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig5}
\end{figure}

\begin{figure}[ht]
\includegraphics{figures/fig6.pdf}
\caption{Predicted instantaneous mortality rates its derivative annual mortality probability. A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship. B,D) Predicted growth\textendash mortality response curves for each species. Errors are $\pm$ 95\% BCI.}
\label{fig6}
\end{figure}

@@

The most complicated function is
\begin{equation}
\lambda_i = \big(\underbrace{\alpha \, \exp\left(-\beta \, X_i \right)}_{\text{Growth-dependent}} +  \underbrace{\gamma}_{\text{Baseline}}\big) \underbrace{\delta_t}_{\text{Census-effect}}.
 \label{add_haz}
\end{equation}
